Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-30 application.tex
waiting for zmax3 max rate fig

One should now model in detail each of the terms on the RHS of
Eq. \ref{real_energy}, linearize them, then solve 
the resulting linear eigenvalue problem. This is beyond 
the scope of the present work. Here, we take the simpler approach of
modeling the linearized RHS of Eq. \ref{real_energy} such that it can 
be represented by a thermal relaxation time $t_c$. 
To do so, we first consider separately perturbations
with radial lengthscales $l\gg l_\mathrm{rad}$ and  
$l\ll l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

@@

However, Eq. \ref{diff_cool_proper} fundamentally changes the linear
problem by introducing higher order derivatives. We defer a proper
analysis of this problem to a follow-up study. Here, we are interested
in order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed 
by assuming vertical derivatives can be neglected in 
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. 
From Eq. \ref{diff_cool_approx} we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}.%\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}

@@

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{\Sigma^2\Omega_k}{\eta\rho^2}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that $\beta$ depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane.

@@

In the inner few AU of the MMSN, however, $\beta$ 
may be smaller or greater than $\beta_\mathrm{crit}$ depending on
$z$ as the disk transitions from an optically thick midplane to an
optically thin atmosphere. 
%Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
%the optically thick midplane, but with increasing height the disk 
%becomes optically thin and $\beta<\beta_\mathrm{crit}$. 
Our analytical criteria is not valid for variable $\beta$, and we must
turn to numerical solutions of the linear problem to determine whether
or not the VSI operates. We do this in the next section.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
for sufficiently large $\khat$. We
illustrate this in Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at 
each radius the value of $\khat$ such  
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI would only
occur on very small radial scales in the inner disk.

@@

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\nu$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\Omega_k$.  In fact, we either only find the fundamental mode, or
that it is the most unstable, except for the appearance of some
surface modes towards the inner disk when considering large
wavenumbers $\khat\gtrsim 50$. However,
surface modes are affected by boundary conditions and disturbances
with large $\khat$ are more likely subject to viscous decay, as
discussed above. The dominance of the fundamental mode is also
observed in hydrodynamic simulations of \cite{stoll14} which treat
radiative cooling properly.

We can therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits. It is necessary to
consider smaller scales with decreasing $r$ as the disk generally becomes
optically thick, and growth times rise rapidly towards the inner disk.

@@

>>>>

One should now model in detail each of the terms on the RHS of
Eq. \ref{real_energy}, linearize them, then solve 
the resulting linear eigenvalue problem. This is beyond 
the scope of the present work. Here, we take the simpler approach of
representing the linearized RHS of Eq. \ref{real_energy} by a thermal
relaxation time $t_c$. To do so, we first consider separately
perturbations with radial lengthscales $l\gg l_\mathrm{rad}$ and  
$l\ll l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

@@

However, Eq. \ref{diff_cool_proper} fundamentally changes the linear
problem by introducing higher order derivatives. We defer a proper
analysis of this problem to a follow-up study. Here, we are interested
in order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed 
by assuming vertical derivatives can be neglected in 
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
\citep[see also][]{barker15} where $\eta=k_\mathrm{rad}/\rho c_v$ is
the diffusion coefficient and 
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. 
From Eq. \ref{diff_cool_approx} we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}.%\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}

@@

\subsection{A model for combined thermal relaxation}\label{toy_relax}
%To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, 
We define the effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{\Sigma^2\Omega_k}{\eta\rho^2}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that $\beta$ depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane.

@@

In the inner few AU of the MMSN, however, $\beta$ 
may be smaller or greater than $\beta_\mathrm{crit}$ depending on
$z$ as the disk transitions from an optically thick midplane to an
optically thin atmosphere. 
%Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
%the optically thick midplane, but with increasing height the disk 
%becomes optically thin and $\beta<\beta_\mathrm{crit}$. 
Our analytical criteria is not valid for non-uniform $\beta$, and we must
turn to numerical solutions of the linear problem to determine whether
or not the VSI operates. We do this in the next section.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$ 
for sufficiently large $\khat$. We illustrate this in 
Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at each radius the value
of $\khat$ such that $\beta = \beta_\mathrm{crit}$, which show that in
the inner disk, say $r\sim 1$AU, the VSI would only occur on very
small radial lengthscales. 
%surface modes can develop for large $\khat$, 
%which have small vertical lengthscales scales. 
%also small vertical scales if they are surface modes

@@

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\nu$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\Omega_k$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes towards the inner disk when considering large 
wavenumbers $\khat\gtrsim 50$. However, surface modes are dependent on
boundary conditions in a vertically isothermal disk \citep{barker15},  
and disturbances with large $\khat$ are more likely subject to viscous
decay, as discussed above. % It is also unclear if surface modes are
% dynamically important given they are concentrated in the disk
% atmosphere where the density is small. 
% The dominance of the fundamental mode is also
% observed in hydrodynamic simulations of \cite{stoll14} which treat
% radiative cooling properly.   unclear if surface modes are
% dynamically important because the there is little mass in the disk at

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits and
$t_\mathrm{grow}$ rises rapidly towards the inner disk. It is
necessary to consider smaller scales with decreasing $r$ as the disk
generally becomes optically thick.

@@

%surface modes may develop for large kx, they are not affected
